\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-06 jasa/paper-jasa.tex
A, B -> \alpha, \beta

Let us now consider the more challenging problem of maximum likelihood
inference from multidimensional data generated by a black-box simulator.
For the sake of the illustration, we will assume 5-dimensional observed samples $\mathbf{x}$ generated
from the following process $p_0$:
    \begin{enumerate}
        \item $\mathbf{z} := (z_0, z_1, z_2, z_3, z_4)$, such that
            $z_0 \sim {\cal N}(\mu=A, \sigma=1)$,
            $z_1 \sim {\cal N}(\mu=B, \sigma=3)$,
            $z_2 \sim {\text{Mixture}}({\cal N}(\mu=-2, \sigma=1), {\cal N}(\mu=2, \sigma=0.5))$,
            $z_3 \sim {\text{Exponential}(\lambda=3)}$ and
            $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
        \item $\mathbf{x} := R  \mathbf{z}$, where $R$ is a fixed
        semi-positive definite $5 \times 5$ matrix defining a fixed projection
        of $\mathbf{z}$ into the observed space.
    \end{enumerate}
Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
our goal is to infer the nominal values $A=1$ and $B=-1$ that
were used for generating this data.

\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $(A=1, B=-1)$.}
    \label{fig:4}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[height=12.5em]{figures/fig5a.pdf}
        \caption{True mean negative log-ratio}
        \label{fig:5a}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[height=12.5em]{figures/fig5b.pdf}
        \caption{Estimated mean negative log-ratio}
        \label{fig:5b}
    \end{subfigure}
    \caption{Inference from parameterized approximated likelihood ratios.
             (\ref{fig:5a}) True mean negative log-ratio for the observed data ${\cal D}$. The red dot corresponds to the nominal values $(A=1, B=-1)$, while the green dot is the maximum likelihood estimator $(\hat A=1.006, \hat B=-1.012)$.
             (\ref{fig:5b}) Estimated mean negative log-ratio, as approximated from the constructed objective by the Gaussian process within the Bayesian optimization procedure. The blue dot is the estimated maximum likelihood estimator $(\hat A=0.956, \hat B=-0.869)$ while the white dots correspond to the points sampled during optimization.}
    \label{fig:5}
\end{figure}

As previously, we make use of Eqn.~\ref{eq:mle_withs} for parameter inference
with approximated likelihood ratios and fix the alternate to $\theta_1:=(A=0,
B=0)$. Following recommendations from Section~\ref{sec:param-clf} to reduce
computational requirements, we train a single parameterized classifier
$s(\mathbf{x};\theta:=(A,B),\theta_1)$; in this case, a 2-layer neural network.
Since the generative model is not expensive, the classifier output is calibrated on-the-fly
with isotonic regression for every candidate parameter pair $\theta:=(A,B)$.
Minimizing the mean negative log of the constructed ratio with Bayesian
optimization~\citep{brochu2010tutorial} eventually leads to the estimated
posterior mean shown in Figure~\ref{fig:5b}, which is close to the true objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations,
the estimated minimizer is found at $(\hat A=0.956, \hat B=-0.869$).
Again, repeated experiments (omitted here) confirm that
the proposed procedure yields an unbiased maximum likelihood estimator.

>>>>

Let us now consider the more challenging problem of maximum likelihood
inference from multidimensional data generated by a black-box simulator.
For the sake of the illustration, we will assume 5-dimensional observed samples $\mathbf{x}$ generated
from the following process $p_0$:
    \begin{enumerate}
        \item $\mathbf{z} := (z_0, z_1, z_2, z_3, z_4)$, such that
            $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
            $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
            $z_2 \sim {\text{Mixture}}({\cal N}(\mu=-2, \sigma=1), {\cal N}(\mu=2, \sigma=0.5))$,
            $z_3 \sim {\text{Exponential}(\lambda=3)}$ and
            $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
        \item $\mathbf{x} := R  \mathbf{z}$, where $R$ is a fixed
        semi-positive definite $5 \times 5$ matrix defining a fixed projection
        of $\mathbf{z}$ into the observed space.
    \end{enumerate}
Given observations ${\cal D}$ represented in Figure~\ref{fig:4},
our goal is to infer the nominal values $\alpha=1$ and $\beta=-1$ that
were used for generating this data.

\begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{figures/fig4.pdf}
    \caption{Observed 5-dimensional data ${\cal D}$ for nominal values $(\alpha=1, \beta=-1)$.}
    \label{fig:4}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[height=12.5em]{figures/fig5a.pdf}
        \caption{True mean negative log-ratio}
        \label{fig:5a}
    \end{subfigure}
    ~
    \begin{subfigure}[b]{0.45\textwidth}
        \includegraphics[height=12.5em]{figures/fig5b.pdf}
        \caption{Estimated mean negative log-ratio}
        \label{fig:5b}
    \end{subfigure}
    \caption{Inference from parameterized approximated likelihood ratios.
             (\ref{fig:5a}) True mean negative log-ratio for the observed data ${\cal D}$. The red dot corresponds
             to the nominal values $(\alpha=1, \beta=-1)$, while the green dot is the maximum likelihood estimator
             $(\hat \alpha=1.006, \hat \beta=-1.012)$.
             (\ref{fig:5b}) Estimated mean negative log-ratio, as approximated from the constructed objective by
             the Gaussian process within the Bayesian optimization procedure. The blue dot is the estimated maximum
             likelihood estimator $(\hat \alpha=0.956, \hat \beta=-0.869)$ while the white dots correspond to
             the points sampled during optimization.}
    \label{fig:5}
\end{figure}

As previously, we make use of Eqn.~\ref{eq:mle_withs} for parameter inference
with approximated likelihood ratios and fix the alternate to
$\theta_1:=(\alpha=0, \beta=0)$. Following recommendations from
Section~\ref{sec:param-clf} to reduce computational requirements, we train a
single parameterized classifier $s(\mathbf{x};\theta:=(\alpha,\beta),\theta_1)$;
in this case, a 2-layer neural network. Since the generative model is not
expensive, the classifier output is calibrated on-the-fly with isotonic
regression for every candidate parameter pair $\theta:=(\alpha,\beta)$.
Minimizing the mean negative log of the constructed ratio with Bayesian
optimization~\citep{brochu2010tutorial} eventually leads to the estimated
posterior mean shown in Figure~\ref{fig:5b}, which is close to the true
objective illustrated in Figure~\ref{fig:5a}. After $50$ iterations, the
estimated minimizer is found at $(\hat \alpha=0.956, \hat \beta=-0.869$). Again,
repeated experiments (omitted here) confirm that the proposed procedure yields
an unbiased maximum likelihood estimator.
